---
title: "Analyzing the Gender Wage Gap"
author: "by Anny-Claude Joseph"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br>

### Background 
The gender wage gap refers to the difference in earnings between females and males in the workforce. The Equal Pay Act of 1963 and the Civil Rights Act of 1964 guarantee protections against discrimination based on traits including age, race, and sex; however, many studies suggest that the gender wage gap has not been eliminated entirely. Let us investigate the gender wage gap using  a dataset based on a portion of the Current Population Survey Uniform Extracts from the Center for Economic and Policy Research (Center, 2019).^[All variable modifications are documented [here](https://github.com/tooderful/MA376/tree/master/LSN%202/Prep%20Files).] For this analysis let us focus on *adult earners* who made $150,000 or less per year. The variables in the dataset are described in the table.
 <br> 
 
<style type="text/css">
.tg  {border-collapse:collapse;border-color:#ccc;border-spacing:0;border-style:solid;border-width:1px;}
.tg td{background-color:#fff;border-color:#ccc;border-style:solid;border-width:0px;color:#333;
  font-family:Arial, sans-serif;font-size:14px;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{background-color:#f0f0f0;border-color:#ccc;border-style:solid;border-width:0px;color:#333;
  font-family:Arial, sans-serif;font-size:14px;font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-fymr{border-color:inherit;font-weight:bold;text-align:left;vertical-align:top}
.tg .tg-btxf{background-color:#f9f9f9;border-color:inherit;text-align:left;vertical-align:top}
.tg .tg-0pky{border-color:inherit;text-align:left;vertical-align:top}
</style>
<table class="tg">
<thead>
  <tr>
    <th class="tg-fymr">Variable</th>
    <th class="tg-fymr"> Description</th>
  </tr>
</thead>
<tbody>
  <tr>
    <td class="tg-btxf">Income</td>
    <td class="tg-btxf">in US dollars</td>
  </tr>
  <tr>
    <td class="tg-0pky">Age</td>
    <td class="tg-0pky">in years</td>
  </tr>
  <tr>
    <td class="tg-btxf">Sex</td>
    <td class="tg-btxf">Male; Female</td>
  </tr>
  <tr>
    <td class="tg-0pky">Race</td>
    <td class="tg-0pky">White; Black; Hispanic; Asian; Other</td>
  </tr>
  <tr>
    <td class="tg-btxf">Education</td>
    <td class="tg-btxf">Less than high school; High school; Some college/Associates; Bachelors; PhD/MD/JD</td>
  </tr>
  <tr>
    <td class="tg-0pky">MaritalStatus</td>
    <td class="tg-0pky">Never married; Married; Separated; Divorced; Widowed</td>
  </tr>
  <tr>
    <td class="tg-btxf">Employed</td>
    <td class="tg-btxf"> Yes ; No</td>
  </tr>
</tbody>
</table>

<br>

```{r, message=FALSE, echo=TRUE}
library('tidyverse')
wage.dat <- read_csv('C:/Users/annyclaude.joseph/Desktop/MA376/CEPR2018.csv')
# head(wage.dat)

summary(wage.dat$Age)        # Verify (adult -> Age >=18)
summary(wage.dat$Income)     # Verify income (earners -> income > 0)

# Create new dataset with individuals who ear $150K or less
wage.dat2 <- wage.dat %>% filter(Income > 0 & Income <= 150000)

``` 
#### Step 1: Ask a research question
What is the research question?
<br>

*Does income for females differ significantly from income for males?*
<br>
<br>

#### Step 2:  Design a study and collect data
Use no more than two sentences to describe the data. Be sure to indicate how many observational units are in the analysis and what values the variables take on. 

```{r, warning=FALSE, message=FALSE}
dim(wage.dat2)
```
<br>

*In this study there were n = 79,394 adult earners (Male/Female) in the United States with incomes of $150,000 or less.*
<br>
<br>

#### Step 3: Explore the data
##### a. Which variable is the explanatory variable? Which is the response variable?
<br>

*Income is the response variable and sex is the explanatory variable.*
<br>

##### b. Summarize income using a histogram and a boxplot. Do the mean and median income differ? Which is the more appropriate measure of central tendency (average)? Why?

```{r, message=FALSE, echo=TRUE}
# Histogram
wage.dat2 %>% ggplot(aes(x=Income)) + geom_histogram(bins=100) + labs(title = 'Distribution of Income')   
# Boxplot
wage.dat2 %>% ggplot(aes(y=Income)) + geom_boxplot()+ coord_flip() + labs(title = 'Distribution of Income')     

library(table1)
table1(~Income, data=wage.dat2)
```
<br>

*The plot of the distribution of income a) has one main peak (unimodal) at around $30,000 and b) is right-skewed. Incomes range between $4 and $150,000 with a mean of $44,200 and median of $37,000. When data are skewed, unusual values are more likely to affect the mean than the median; thus, it is common to use the median as a measure of the average for these situations.*
<br>

##### c. Use boxplots to plot the conditional distributions of income for males and females. Consider whether there appears to be an association between the two variables. Can we conclude that there is a wage (income) gap? Explain.

```{r, message=FALSE, echo=TRUE}
wage.dat2 %>% ggplot(aes(y=Income,x=Sex)) + geom_boxplot() + coord_flip() + labs(title = 'Income Distributions Conditional on Sex')
table1(~Income|Sex,  data = wage.dat2)  
```
<br>

*Overall, there is some evidence of a wage (income) gap. The conditional distributions of income were substantially different for males and females. This suggests that there is an association between income and sex. Median (and mean) income for males was more than $12,000 higher than for females. Also, there was more variability in income for males ($34,500 vs $29,200). *
<br>

##### d. Identify a potential confounding variable and explain how it may be confounding in this study. 
<br>

*Education is a third variable that potentially explains variation in income, but it might also be  associated with sex.*
<br> 

##### e. Use an appropriate plot(s) to explore the association between income and the third variable. Is there evidence of a relationship between these two variables? Explain.

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# wage.dat2 %>% ggplot(aes(y=Income,x=Education))+geom_boxplot()+coord_flip()

# Reorder education levels so trend is more obvious
wage.dat2$Education <- factor(wage.dat2$Education, levels = c('Less than HS','HS Graduate',
                                                    'Some College/Associates', 'Bachelors',
                                                    'Masters', 'PhD/MD/JD'))
wage.dat2 %>% ggplot(aes(y=Income,x=Education))+geom_boxplot()+coord_flip() + labs(title = 'Income Distributions Conditional on Level of Education')
``` 
<br>

*There is evidence of an association between education and income. On average individuals at higher education levels tended to have higher incomes.*
<br>

##### f. Describe the relationship between sex and third variable.  

```{r, message=FALSE, warning=FALSE, echo=TRUE}
table1(~Sex|Education, data = wage.dat2)
```
<br>

*The proportion of males and females were near parity at the highest level of education (PhD/MD/JD). A higher proportion of males tended to have lower education levels, whereas a larger proportion of females tended to have some college/Associates degrees, Bachelor's degrees, and Masters degrees. *
<br>

##### g. Create boxplots for income distributions conditional on the level of the third variable  for males and females. Do the plots suggest that the third variable is confounding? Explain.

```{r, message=FALSE, warning=FALSE, echo=TRUE}
wage.dat2 %>% ggplot(aes(y=Income,x=Education,fill=Sex)) + geom_boxplot()  +coord_flip() + labs(title = 'Income Distributions Conditional on Education \n for Males vs Females')
# table1(~Income|Sex*Education,  data = wage.dat2)
``` 
<br>

*Average income for females was lower than for males across all education levels. Median income was noticeably larger for males in education levels with higher proportion of females. This suggests that education is not a confounding variable in the relationship between sex and income.*
<br>

##### h. Does sex explain all the variation we observe in income? Explain using a Sources of Variation Diagram:^[A convenient table generator can be found [here](https://www.tablesgenerator.com/latex_tables).]
<br>

<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-0pky{border-color:inherit;text-align:left;vertical-align:top}
.tg .tg-fymr{border-color:inherit;font-weight:bold;text-align:left;vertical-align:top}
.tg .tg-3qbk{border-color:inherit;color:#3166ff;text-align:left;vertical-align:top}
.tg .tg-6tgj{border-color:inherit;color:#f56b00;text-align:left;vertical-align:top}
.tg .tg-tu1s{border-color:inherit;color:#32cb00;text-align:left;vertical-align:top}
.tg .tg-f8tv{border-color:inherit;font-style:italic;text-align:left;vertical-align:top}
</style>
<table class="tg">
<thead>
  <tr>
    <th class="tg-6tgj">Observed variation in:</th>
    <th class="tg-tu1s">Sources of explained variation: </th>
    <th class="tg-3qbk">Sources of unexplained variation:</th>
  </tr>
</thead>
<tbody>
  <tr>
    <td class="tg-0pky">Income</td>
    <td class="tg-0pky">Sex</td>
    <td class="tg-0pky">Type of job</td>
  </tr>
  
  <tr>
    <td class="tg-0pky"></td>
    <td class="tg-0pky">Education</td>
    <td class="tg-0pky">Race</td>
  </tr>
  <tr>
    <td class="tg-f8tv">Inclusion criteria:</td>
    <td class="tg-0pky"></td>
    <td class="tg-0pky">Marital Status</td>

  </tr>
  <tr>
    <td class="tg-0pky">U.S. adults who earn $150k or less</td>
    <td class="tg-0pky"></td>
    <td class="tg-0pky">Unknown</td>
  </tr>
</tbody>
</table>
<br>
<br>

#### Step 4: Draw inferences beyond the data
Name one theory-based approach we could use to determine if there was a statistically significant difference in incomes between the sexes.
<br> 

*Two-sample t-test could be used to determine whether income for females differs significantly from income for males.*
<br>
<br>

#### Step 5: Formulate conclusions
Summarize the conclusions we could draw from this study. How far can we generalize our conclusions? Could we draw a cause-effect conclusion if we observed statistically significant differences in Step 4? Why or why not?
<br>

*There appears to be an association between income and sex; however, we would not be able to make a cause-effect conclusion because this was an observational study and not a randomized experiment. Although the data suggest that education may be related to income and sex separately, the education variable does not distort the conclusions we would draw about the relationship between income and sex.*
<br>
<br>

#### Step 6: Look back and ahead
Identify limitations to our approach. What could we do to address them in a future study.

*A major limitation of this study is that it does not allow us to account for other sources of variation like type of job and race. In future work we could create statistical models that allow us to account for these variables.*
<br>
<br>

### Further Investigation
Click [here](https://army-math-west-point.shinyapps.io/CEPRVisualizer/) 
for an interactive, custom-designed applet that can be used to explore this dataset further.
<br>
<br>

### References
Center for Economic and Policy Research. March CPS Uniform Extracts, Version 1.1. Washington, DC.
https://ceprdata.org/cps-uniform-data-extracts/cps-outgoing-rotation-group/cps-org-data/. (2019). Accessed: 2020-08-12.




