---
title: "Lesson 6"
author: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library('tidyverse')

wage.dat <- read_csv('/CEPR2018_3.csv')
# head(wage.dat)


# Insert your code here
``` 
### Background 
We continue to work with the dataset based on a portion of the Current Population Survey Uniform Extracts from the Center for Economic and Policy Research (Center, 2019).^[All variable modifications are documented [here](https://github.com/tooderful/MA376/blob/master/LSN%206/Prep%20Files/CEPR_2018_Data3.R).] For this analysis we will focus on a random sample of 1,400 *adult earners with Bachelors degrees who made $150,000 or less per year*. 


#### Step 1: Ask a research question.  

What is the research question?

*Your response goes here* 
\vspace{.5in}


#### Step 2:  Design a study and collect data.

Use no more than three sentences to describe the data. Be sure to indicate how many observational units are in the analysis and what values the variables take on.

*Your response goes here* 
\vspace{1in}


#### Step 3: Explore the data.
Use boxplots to plot the conditional distributions of income between the races. Consider whether there appears to be an association between the two variables.

*Your response goes here* 
\vspace{.75in}

#### Step 4: Draw inferences beyond the data.

The difference in income is practically meaningful, but is it beyond what we would expect to observe by chance? That is, is the difference in income statistically significant at the $\small \alpha$ = 0.05 significance level? Write the null and alternative hypothesis in words and symbols.

*Your response goes here* 
\vspace{1.5in}

We can use two approaches to determine statistical significance: the 3S strategy or the theory-based approach.

***A. 3S Strategy***

Use the [Comparing Groups applet](http://www.rossmanchance.com/applets/ComparingGroups.html). Make sure the explanatory and response variables are ordered to match the button above the data and press 'Use Data'. Check the box to 'Show Groups'. Check the 'Show Shuffle Options' box and select the 'Plot' radio button. Use at least 1000 shuffles.  

a) We can use $\small R^2$ as the sample statistic. What is the value of $\small R^2$? Explain what the value means.

*Your response goes here* 
\vspace{.5in}

b) What is the standardized statistic?

*Your response goes here* 
\vspace{.5in}

c) Is there evidence of a statistically significant difference in mean income between the races? 

*Your response goes here* 
\vspace{.75in}

***B. Theory-based approach*** 

a. What theory-based approach could we use to determine if there was a statistically significant difference in incomes between the racial groups?

*Your response goes here* 
\vspace{.5in}

 
b. Are the validity conditions for this theory-based approach satisfied?

*Your response goes here* 
\vspace{1in}

 
c. Are the results for the theory-based approach consistent with those from the 3S Strategy? 

*Your response goes here* 
\vspace{.25in}  
 
 
#### Step 5: Formulate conclusions.
Summarize the conclusions we could draw from this study. How far can we generalize our conclusions?

*Your response goes here*
\vspace{1in} 

 
 
#### Step 6: Look back and ahead.
Identify limitations to our approach. What could we do to address them in a future study?

*Your response goes here*
\vspace{0.85in} 


 

### References
Center for Economic and Policy Research. March CPS Uniform Extracts, Version 1.1. Washington, DC.
https://ceprdata.org/cps-uniform-data-extracts/cps-outgoing-rotation-group/cps-org-data/. (2019). Accessed: 2020-08-12.




