---
title: "Analyzing the Gender Wage Gap (continued)"
author: "by Anny-Claude Joseph"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
---

```{r setup, message=FALSE, echo=TRUE, include=FALSE, warning=FALSE}
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color, 
      x)
  } else x
}
knitr::opts_chunk$set(echo = TRUE)
```
<br>

### Background 
Let us revisit the investigation of the gender wage gap started in Lesson 2. The dataset we are using is based on a portion of the Current Population Survey Uniform Extracts from the Center for Economic and Policy Research (Center, 2019).^[All variable modifications are documented [here](https://github.com/tooderful/MA376/blob/master/LSN%205/Prep%20Files/CEPR_2018_Data2.R).] For this analysis we will focus on a random sample of 1,400 *adult earners with Bachelors degrees who made $150,000 or less per year*. 

```{r, message=FALSE, echo=TRUE}
library('tidyverse')
wage.dat <- read_csv('C:/Users/annyclaude.joseph/Desktop/MA376/CEPR2018_2.csv')
``` 

<br>

#### Step 1: Ask a research question
`r colorize("*Does income for females differ significantly from income for males?*", "blue")`

<br>

#### Step 2:  Design a study and collect data
`r colorize("*In this study there were 700 adult males and 700 adult females with Bachelors degrees who earned $150,000 or less.*", "blue")`

```{r, message=FALSE, echo=TRUE}
library(table1)
table1(~Sex, data = wage.dat)
``` 

<br>

#### Step 3: Explore the data
`r colorize("*The conditional distributions of income were substantially different for males and females. Mean income for males was more than $18,000 higher than for females.*", "blue")`
  

```{r, message=FALSE, echo=TRUE, out.width = '60%',fig.align = "center"}
wage.dat %>% ggplot(aes(y=Income,x=Sex, fill=Sex)) + geom_boxplot() + labs(title = 'Income Distributions Conditional on Sex')
table1(~Income|Sex, render='Mean (SD)', data = wage.dat)
```

<br>

#### Step 4: Draw inferences beyond the data
The difference is practically meaningful, but is it beyond what we would expect to observe by chance? That is, is the difference in income statistically significant at the $\small \alpha$ = 0.05 significance level? Write the null and alternative hypothesis in words and symbols.

`r colorize("*The null hypothesis is that there is no difference in mean income between males and females with Bachelors degrees in the United States. The alternative hypothesis is that males earn more than females on average.*", "blue")`
$$\small  H_0: \mu_{\text{males}} - \mu_{\text{females}} = 0 \\
\small H_A: \mu_{\text{males}} - \mu_{\text{females}} > 0  $$


We can use two approaches to determine statistical significance: the 3S strategy or the theory-based approach.

<p style="text-align: center;"> `r colorize("**A. 3S Strategy**", "Red")`</p>

Use the [Comparing Groups applet](http://www.rossmanchance.com/applets/ComparingGroups.html). Make sure the explanatory and response variables are ordered to match the button above the data and press 'Use Data'. Check the box to 'Show Groups'. Check the 'Show Shuffle Options' box and select the 'Plot' radio button. Use at least 1000 shuffles.  

a) What is the sample statistic (difference in means)?
 $\bar{x}_{\text{males}} - \bar{x}_{\text{females}} =$ $18,687.52

b) What is the standardized statistic?
`r colorize("*t-statistic = 10.2*", "Blue")`

c) Is there evidence of a statistically significant difference in the mean income for males and females? `r colorize("*Given the p-value < 0.001, we have very strong evidence against the null hypothesis at the 0.05 significance level.*", "Blue")`

d) Interpret the 95% confidence interval.
`r colorize("*We can say with 95% certainty that males with Bachelors degrees earn approximately $15,093 to $22,282 more than females with Bachelors degrees.*", "Blue")`


<p style="text-align: center;"> `r colorize("**B. Theory-based approach**", "Red")`</p>


a. What theory-based approach could we use to determine if there was a statistically significant difference in incomes between the sexes?

`r colorize("*A two-sample t-test could be used to determine whether income for females differs significantly from income for males.*", "blue")`


```{r, message=FALSE, echo=TRUE}
male <- wage.dat %>% filter(Sex == 'M') 
female <- wage.dat %>% filter(Sex == 'F') 

t.test(male$Income, female$Income, alternative='greater', #var.equal = T,
       mu=0, conf.level=0.95)
``` 

b. Are the validity conditions for this theory-based approach satisfied?

`r colorize("*1) The samples (males vs females) are independent of each other; 2) the SD are roughly equal (SD male = $36,400, SD female = $32,100); and 3) both samples have 700 individuals though the distribution of income for females is slightly right skewed.*", "blue")`

```{r, message=FALSE, echo=TRUE, out.width = '50%',fig.align = "center"}
male %>% ggplot(aes(x=Income)) + geom_histogram()
female %>% ggplot(aes(x=Income)) + geom_histogram()
``` 

c. Are the results for the theory-based approach consistent with those from the 3S Strategy? `r colorize("*1) The results are consistent. We have strong evidence against the null hypothesis despite the moderate violation in the validity conditions.*", "blue")`

<br>
 
#### Step 5: Formulate conclusions
Summarize the conclusions we could draw from this study. How far can we generalize our conclusions?

`r colorize("*We reject the null hypothesis of no difference (p-value < 0.001). The data suggest  that males with Bachelors degrees earn more on average than females with Bachelors degrees. Specifically, we can say with 95% certainty that males earn approximately $15,093 to $22,282 more than females with Bachelors degrees. We would be hesitant to generalize our conclusions to other educational levels.*", "blue")`

<br>
 
#### Step 6: Look back and ahead
Identify limitations to our approach. What could we do to address them in a future study?

`r colorize("*A major limitation of this study is that it does not allow us to account for other sources of variation like type of job and race. In future work we could create statistical models that allow us to account for these variables.*", "blue")`


<br>
<br>
<br>
<br>

### References
Center for Economic and Policy Research. March CPS Uniform Extracts, Version 1.1. Washington, DC.
https://ceprdata.org/cps-uniform-data-extracts/cps-outgoing-rotation-group/cps-org-data/. (2019). Accessed: 2020-08-12.




