---
title: "The Gender Wage Gap"
author: "by Anny-Claude Joseph"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br>

### Background 
Under the Equal Pay Act of 1963, employers are obliged to pay men and women equally for jobs requiring the same skill, effort, and responsibility. The Civil Rights Act of 1964 also guarantees protections against discrimination based on traits including age, race, and sex. More than 50 years after these laws were enacted, the wage gap between the sexes has not been eliminated entirely and several highly publicized lawsuits suggest that it remains a controversial topic. One notable example is the 2019 lawsuit filed by the U.S. Women's National Soccer Team against the U.S. Soccer Federation in which the women's team claimed they were underpaid compared to their male counterpart. 

Let us examine public sentiment on this topic using  a dataset based on a portion of the Pew Research Center's American Trends Panel (Pew Research Center, 2019).^[All variable modifications are documented [here](https://github.com/tooderful/MA376/tree/master/LSN%201/Prep%20Files). Additional information about the Pew study can be found in the ATP W49 methodology pdf.] The variables in the dataset are described in the table.

<br>



<br>

```{r, message=FALSE, echo=TRUE}
dat <- read_csv('C:/Users/annyclaude.joseph/Desktop/MA376/CEPR2018.csv')

ggplot(dat,aes(x=Income))+geom_histogram(bins=100)
ggplot(dat,aes(y=Income))+geom_boxplot()+coord_flip()

# Remove individuals with no income  and restrict income to <= 150K
dat <- dat %>% filter(Income >0 & Income <= 150000)   

ggplot(dat,aes(y=Income,x=Sex))+geom_boxplot()+coord_flip()
table1(~Income|Sex,  data = dat)  
 
ggplot(dat,aes(y=Income,x=Education,fill=Sex))+geom_boxplot()+coord_flip()
table1(~Income|Sex*Education,  data = dat)
# render.continuous=c("Mean (SD)",.="Median [Min, Q1, Q3, Max]")


``` 
#### Step 1: Ask a research question
Write a research question involving two variables in the dataset.

*Is there an association between educational attainment and belief that law enforcement would use FRT responsibly?*

<br>
<br>

#### Step 2:  Design a study and collect data
Use a few sentences to describe the data. Be sure to indicate how many subjects are in the analysis and what values the variables take on. 

*Your answer *

```{r, warning=FALSE, message=FALSE}
library('table1')
table1(~TrustPolice+Education, data = dat)
# table(dat$TrustPolice)
# table(dat$Education)
```

<br>
<br>

#### Step 3: Explore the data
##### a. Create a bar graph of the response variable. Describe the plot.

*More than 2,000 adults agreed (Yes/Somewhat) that law enforcement could be trusted to use FRT responsibly while a little over 1,000 expressed more skepticism (Not too much/No)*.
```{r, warning=FALSE}
# a. Bar graph   
dat %>% ggplot(aes(TrustPolice))+ geom_histogram(stat='count')
```

<br>

##### b. Create a contingency table and mosaic plot to summarize the response and explanatory variables. Does the plot suggest an association exists? Explain.

*The plot shows that a higher proportion of individuals did not have a college education (59.1%), and higher proportion of individuals trusted law enforcement (68.1%). The percentage of individuals who trusted law enforcement was higher for non-college educated than college educated individuals (70.2% vs 65.2%).*

```{r, warning=FALSE}
# b. Contingency table 
table1(~TrustPolice|Education, data = dat)

# Mosaic plot
# install.package(ggmosaic)
library('ggmosaic')
dat %>% ggplot()+
  geom_mosaic(aes(x=product(TrustPolice,Education),fill=Education))+
  labs(title='Public sentiment on law enforcement use of facial recognition technology', y='Trust Law Enforcement', x='Education')
```

<br>

##### c. Identify a potential confounding variable and explain how it may be confounding in this study. 

*Race is a potential confounder. Sentiments about law enforcement use of FRT and educational attainment may vary by race*.
```{r}
table1(~TrustPolice+Education|Race, data =dat)
```

<br>

##### d. Create another mosaic plot that includes the third variable. Does the plot suggest that the third variable is confounding? Explain.

*Race impacts the association between educational attainment and belief that law enforcement would use FRT responsibly. A larger proportion of Whites trusted law enforcement regardless of educational attainment. However, for non-college educated Blacks, the percentage of individuals who trusted law enforcement was higher than those who were more skeptical (55.8% vs 44.2%); whereas for college educated Blacks, the percentage of individuals who trusted law enforcement was lower (46.3% vs 53.7%).*
```{r, warning=FALSE}
# d. Mosaic plot that includes the third variable
dat %>% ggplot()+ 
  geom_mosaic(aes(x=product(TrustPolice,Education),fill=Education)) +
  labs(title='Public sentiment on law enforcement use of facial recognition technology \n for Blacks vs Whites', y='Trust Law Enforcement', x='Education')+
  facet_grid(Race~.)
table1(~TrustPolice|Race*Education, data = dat)
```

<br>

##### e. Construct the Sources of Variation Diagram to summarize what we have learned so far:^[A convenient table generator can be found [here](https://www.tablesgenerator.com/latex_tables).]
<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-0pky{border-color:inherit;text-align:left;vertical-align:top}
.tg .tg-fymr{border-color:inherit;font-weight:bold;text-align:left;vertical-align:top}
</style>
<table class="tg">
<thead>
  <tr>
    <th class="tg-fymr">Observed variation in:</th>
    <th class="tg-fymr">Sources of explained variation: </th>
    <th class="tg-fymr">Sources of unexplained variation:</th>
  </tr>
</thead>
<tbody>
  <tr>
    <td class="tg-0pky">Trust law enforcement</td>
    <td class="tg-0pky">Education</td>
    <td class="tg-0pky">Age</td>
  </tr>
  
  <tr>
    <td class="tg-0pky"></td>
    <td class="tg-0pky">Race</td>
    <td class="tg-0pky">Socioeconomic status</td>
  </tr>
  <tr>
    <td class="tg-fymr">Inclusion criteria:</td>
    <td class="tg-0pky"></td>
    <td class="tg-0pky">Political affiliation</td>

  </tr>
  <tr>
    <td class="tg-0pky">U.S. adults </td>
    <td class="tg-0pky"></td>
    <td class="tg-0pky">Unknown</td>
  </tr>
</tbody>
</table>

<br>
<br>

#### Step 4: Draw inferences beyond the data
Name one theory-based approach we could use to determine if there was a statistically significant difference between the levels of the explanatory variable.

*Your answer *

<br>
<br>

#### Step 5: Formulate conclusions
Summarize the conclusions you would draw from this study. How far can we generalize our conclusions? Could we draw a cause-effect conclusion if we observed statistically significant differences in Step 4? Why or why not?

*Your answer *

<br>
<br>

#### Step 6: Look back and ahead
Identify some limitations to our approach and what we could do to address them in a future study.

*Your answer *

<br>

### Further Investigation
Work with a partner to investigate another response variable. Overall, are the conclusions consistent with this analysis?

<br>
<br>

### References
Pew Research Center. American Trends Panel Wave 49. https://www.pewresearch.org/internet/dataset/american-trends-panel-wave-49/ (2019). Accessed: 2020-08-12.
 

